---
title: "Prepare input data for clonevol"
author: 'Antonia Chroni <chronia@chop.edu> for D3B'
date: "2023"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
---

# Background

[ClonEvol](https://github.com/hdng/clonevol) is a package for clonal ordering and clonal evolution visualization. It uses the clustering of heterozygous variants identified using other tools as input to infer consensus clonal evolution trees and estimate the cancer cell fraction (also called clonal frequency) of the clones in individual samples. ClonEvol can deal with statistical uncertainty and error in sequencing and data analysis that may distort the cellular prevalence estimate of individual variants.


# Set up

```{r load-library}
suppressPackageStartupMessages({
  library(tidyverse)
})
```

# Directories and File Inputs/Outputs

```{r set-dir-and-file-names}
# Detect the ".git" folder -- this will be in the project root directory
# Use this as the root directory to ensure proper sourcing of functions 
# no matter where this is called from
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
analysis_dir <- file.path(root_dir, "analyses", "tumor-clone-inference")
results_dir <- file.path(analysis_dir, "results")
input_dir <- file.path(results_dir, "fastclone_output")
#goi_input_dir <- file.path(root_dir, "analyses", "tmb-vaf-longitudinal", "results")


# Input files
nautilus_dec_file <- file.path(results_dir, "nautilus_dec.tsv") 
fastclone_output_file <- file.path(input_dir, "PT_KZ56XHJT/scores.csv") 
#goi_list_file <- file.path(goi_input_dir, "df_list_common_genes.tsv") 
signif_genes_file <- file.path(results_dir, "signif_genes.tsv") 

# File path to input directory
clonevol_input_dir <-
  file.path(analysis_dir, "results", "clonevol-input")
if (!dir.exists(clonevol_input_dir)) {
  dir.create(clonevol_input_dir)
}
```

# Load and process data

```{r load-process-inputs}
# It is important to order timepoints 
# so plots and numbers of clusters/clones will appear in the correct order for plots

# Define timepoints
timepoints = c("Diagnosis", "Progressive", "Recurrence", "Deceased", "Second Malignancy", "Unavailable")

patient_df <- readr::read_tsv(nautilus_dec_file, guess_max = 100000, show_col_types = FALSE) %>%
  filter(Kids_First_Participant_ID == "PT_KZ56XHJT") %>% 
  select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, cg_id, tumor_descriptor,
         sample_id, t_ref_count, t_alt_count, mutation_id,
         tumor_fraction, VAF, Hugo_Symbol, CCF) %>% 
  mutate(tumor_descriptor = factor(tumor_descriptor),
         tumor_descriptor = fct_relevel(tumor_descriptor, timepoints)) %>% 
  arrange(tumor_descriptor)

# Read and process cluster file
cluster_df <- read_csv(fastclone_output_file) 

colnames(cluster_df)[1] <- "mutation_id" 
names(cluster_df)

# Count number of columns
count_col <- ncol(cluster_df)


pattern = c(":*:")

cluster_df <- cluster_df %>% 
  mutate(mutation_id = str_replace(mutation_id, "*:", ""))





  
  
  # change names to match input requirements
  dplyr::rename("cluster" = "cluster_id") %>% 
  select(cluster, sample_id) %>% 
  as.data.frame()

clonevol_df <- patient_df %>% 
  select(Kids_First_Participant_ID, tumor_descriptor, sample_id, Hugo_Symbol, t_ref_count, t_alt_count,
         VAF, CCF) %>% 
  left_join(cluster_df) %>% 
  # change names to match input requirements
  dplyr::rename("ref.count" = "t_ref_count", 
                "var.count" = "t_alt_count",
                "vaf" = "VAF",
                "ccf" = "CCF") %>% 
  unique() %>% 
  mutate_at(c(5:9), as.numeric) %>% 

  # pyclone-vi starts numbering of clones from `0`
  # this will cause an error for clonevol when trying to subset
  # in the `plot.variant.clusters`
  # In addition, `infer.clonal.models` function doesn't accept 0 for clustering
  dplyr::mutate(cluster = case_when(grepl("Deceased", tumor_descriptor) ~ "3",
                                    grepl("Recurrence", tumor_descriptor) ~ "2",
                                    TRUE ~ "1"),
                # cluster = case_when(grepl("0", cluster) ~ "1"),
                ccf = (ccf*100),
                vaf = (vaf*100)#,
                #unique_id = paste(sample_id, ref.count, sep = ":")
                )  


# to make a loop for each `Kids_First_Participant_ID`
# for now it's one sample
# make long table

df <- clonevol_df %>%
  # ref.count
  mutate(id = paste(tumor_descriptor, "ref.count", sep = ".")) %>% #sample_id

  # USING pivit_wider for multiple columns was intrudicng a bug so I split it up instead
  #pivot_wider(names_from = "sample_id", values_from = c("ref.count", "var.count", "vaf", "ccf")) %>% #values_fn = list
  pivot_wider(names_from = "id", values_from = c("ref.count"), values_fill = 0) %>% #values_fn = list
  
  # var.count
  left_join(clonevol_df, relationship = "many-to-many") %>% 
  select(-c(ref.count)) %>% 
  mutate(id = paste(tumor_descriptor, "var.count", sep = ".")) %>% 
  pivot_wider(names_from = "id", values_from = c("var.count"), values_fill = 0) %>% #values_fn = list

  # vaf
  left_join(clonevol_df, relationship = "many-to-many") %>% 
  select(-c(ref.count, var.count)) %>% 
  mutate(id = paste(tumor_descriptor, "vaf", sep = ".")) %>% 
  pivot_wider(names_from = "id", values_from = c("vaf"), values_fill = 0) %>% #values_fn = list
  
  # ccf
  left_join(clonevol_df, relationship = "many-to-many") %>% 
  select(-c(ref.count, var.count, vaf)) %>% 
  mutate(id = paste(tumor_descriptor, "ccf", sep = ".")) %>% 
  unique() %>% 
  pivot_wider(names_from = "id", values_from = c("ccf"), values_fill = 0) %>% #values_fn = list
  
  as.data.frame() %>% 
  select(-c(Kids_First_Participant_ID, tumor_descriptor)) 
  
# Count number of columns
count_col <- ncol(df)

signif_genes_df <- readr::read_tsv(signif_genes_file, guess_max = 100000, show_col_types = FALSE) %>% 
  select(gene_name) %>% 
  dplyr::rename("gene" = "gene_name") %>% 
  dplyr::mutate(is.driver = "TRUE")

# Make input for clonevol
clonevol_input_df <- df %>% 
  mutate_at(c(3:count_col), as.numeric) %>%
  dplyr::rename("gene" = "Hugo_Symbol") %>% 
  #mutate(id = paste(sample_id, Hugo_Symbol, sep = "_")) %>% 
  #select(-c(Hugo_Symbol)) %>% 
  #left_join(goi_df) %>% 
  left_join(signif_genes_df) %>%
  dplyr::mutate(is.driver = replace_na(is.driver, "FALSE")) %>% 
  #dplyr::rename("gene" = "gene_protein", 
  #              # we will use sym info as driver info for now
  #             "is.driver" = "sym") %>%
  select(-c(sample_id)) %>%
  relocate(c(gene, is.driver), .after = Deceased.ccf) %>% 
  unique %>%
  filter(!is.na(is.driver)) %>% 
  replace(is.na(.), "-")
  #filter(!is.na(is.driver)) %>% 
  #drop_na() 


# Convert labels to display on the plot only the ones in the goi list
clonevol_input_df$gene <- ifelse(clonevol_input_df$is.driver == TRUE, clonevol_input_df$gene, "-")


clonevol_input_df <- clonevol_input_df %>%
  #mutate(gene = paste(gene, Diagnosis.ref.count, sep = "_")) %>% 
  write_tsv(file.path(clonevol_input_dir, "clonevol_input.tsv"))
```


```{r}
sessionInfo()
```


