---
title: "Create TMB stacked barplots of tumors across multiple timepoints for thr autopsy samples in the PBTA Cohort"
author: "Antonia Chroni <chronia@chop.edu> and Jo Lynne Rokita <rokita@chop.edu> for D3B"
date: "2023"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
---

#### Tumor evolution project 

### Data used 
title: "Create TMB stacked barplots of tumors across multiple timepoints for thr autopsy samples in the PBTA Cohort"

# Set up
```{r load-library}
suppressPackageStartupMessages({
  library(tidyverse)
  library(reshape2) 
})
```

# Directories and File Inputs/Outputs
```{r set-dir-and-file-names}
# Detect the ".git" folder. This will be in the project root directory.
# Use this as the root directory to ensure proper sourcing of functions
# no matter where this is called from.
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
scratch_dir <- file.path(root_dir, "scratch")
analysis_dir <- file.path(root_dir, "analyses", "tmb-vaf-longitudinal") 
input_dir <- file.path(analysis_dir, "input")

# Input files
tmb_autopsy_file <- file.path(scratch_dir, "tmb_autopsy.tsv")

# File path to plots directory
plots_dir <-
  file.path(analysis_dir, "plots")
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}

source(paste0(analysis_dir, "/util/function-create-stacked-barplot.R"))
source(paste0(analysis_dir, "/util/function-create-barplot.R"))
source(paste0(root_dir, "/figures/theme.R"))
```

# Read in data and process
```{r read_input_files}
# Read tmb_autopsy file generated from step `01-process-data.Rmd`
tmb_autopsy <- readr::read_tsv(tmb_autopsy_file, guess_max = 100000, show_col_types = FALSE)
```

# Stacked barplots for each Patient case
```{r create-stacked-barplot}
# Create one stacked barplot with all samples together
ylim <- 13000
fname <- paste0(plots_dir, "/", "TMB-stacked-barplot-All-Samples.pdf")
print(fname)
p <- create_stacked_barplot(tmb = tmb_autopsy, ylim = ylim)
pdf(file = fname, width = 10, height = 8)
print(p)
dev.off()
```

```{r create-stacked-barplot-exclude}
# Exclude 2 samples with high number of mutations to make plot more readable
# We will exclude PT_3CHB9PK5 - Diagnosis model AND PT_6N825561 - Progessive model
  # This is because the samples contain > 13000 and >2000 mutations, respectively
  # making it hard to adjust y axis and read the plot
ylim <- 400
tmb_autopsy_filter <- tmb_autopsy %>%
  filter(!Kids_First_Participant_ID %in% c("PT_3CHB9PK5", "PT_6N825561"))
fname_exclude <- paste0(plots_dir, "/", "TMB-stacked-barplot-All-Samples-exclude.pdf")
print(fname_exclude)
p <- create_stacked_barplot(tmb = tmb_autopsy, ylim)
pdf(file = fname_exclude, width = 10, height = 8)
print(p)
dev.off()
```

# Stacked barplots for each Patient case and cancer type
```{r create-stacked-barplot}
# Create one stacked barplot with all samples together
ylim <- 13000
fname <- paste0(plots_dir, "/", "TMB-stacked-barplot-All-Samples-cancer-type.pdf")
print(fname)
p <- create_stacked_barplot_cancer_type(tmb = tmb_autopsy, ylim = ylim)
pdf(file = fname, width = 15, height = 10)
print(p)
dev.off()
```

```{r create-stacked-barplot}
# Create one stacked barplot with all samples together
ylim <- 400
fname <- paste0(plots_dir, "/", "TMB-stacked-barplot-All-Samples-exclude-cancer-type.pdf")
print(fname)
p <- create_stacked_barplot_cancer_type(tmb = tmb_autopsy, ylim = ylim)
pdf(file = fname, width = 15, height = 10)
print(p)
dev.off()
```

Since we have multiple biospecimen samples per timepoint and patient case, let's plot these separetely.

# Let's create stacked-barplots per patient sample and split by biospecimen sample
```{r create-stacked-barplot-sample}
samples <- unique(as.character(tmb_autopsy$Kids_First_Participant_ID))
print(samples)

samples_include <- samples[!samples %in% c("PT_3CHB9PK5", "PT_6N825561")]

for (i in seq_along(samples_include)) {
  print(i)
  tmb_sub <- tmb_autopsy %>%
    filter(Kids_First_Participant_ID == samples_include[i])
  ylim <- 100 # to scale yaxis
  
  # Run function
  fname <- paste0(plots_dir, "/", samples_include[i], "-TMB-barplot.pdf")
  print(fname)
  p <- create_barplot_sample(tmb = tmb_sub,
                             sid = samples_include[i],
                             ylim = ylim)
  pdf(file = fname, width = 5, height = 4)
  print(p)
  dev.off()
}

samples_exclude <- samples[samples %in% c("PT_3CHB9PK5", "PT_6N825561")]

for (i in seq_along(samples_exclude)) {
  print(i)
  tmb_sub <- tmb_autopsy %>%
    filter(Kids_First_Participant_ID == samples_exclude[i])
  ylim <- 13000 # to scale yaxis
  
  # Run function
  fname <- paste0(plots_dir, "/", samples_exclude[i], "-TMB-barplot.pdf")
  print(fname)
  p <- create_barplot_sample(tmb = tmb_sub,
                             sid = samples_exclude[i],
                             ylim = ylim)
  pdf(file = fname, width = 5, height = 4)
  print(p)
  dev.off() 
}
```


```{r echo=TRUE}
sessionInfo()
```

