---
title: "Classification of Variants across paired longitudinal samples in the PBTA Cohort"
author: 'Antonia Chroni <chronia@chop.edu> and Jo Lynne Rokita <rokita@chop.edu> for D3B'
date: "2023"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
---

# Set up
```{r load-library}
suppressPackageStartupMessages({
  library(tidyverse)
})
```

## Directories and File Inputs/Outputs
```{r set-dir-and-file-names}
# Detect the ".git" folder -- this will be in the project root directory
# Use this as the root directory to ensure proper sourcing of functions 
# no matter where this is called from
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
analysis_dir <- file.path(root_dir, "analyses", "tmb-vaf-longitudinal")
results_dir <- file.path(analysis_dir, "results")
input_dir <- file.path(analysis_dir, "input")
files_dir <- file.path(root_dir, "analyses", "sample-distribution-analysis", "results")

# Input files
genomic_paired_file <- file.path(files_dir, "list-genomic-paired.tsv") 
tmb_vaf_file <- file.path(results_dir, "tmb_vaf_genomic.tsv")
palette_file <- file.path(root_dir, "figures", "palettes", "oncoprint_color_palette.tsv")

# File path to plot directory
plots_dir <-
  file.path(analysis_dir, "plots")
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}

source(paste0(root_dir, "/figures/scripts/theme.R"))
```

## Read in data and process
```{r load-process-inputs}
tmb_vaf_df <- readr::read_tsv(tmb_vaf_file, guess_max = 100000, show_col_types = FALSE) %>% 
  filter(!tmb >= 10) %>% 
  select(Kids_First_Biospecimen_ID, Variant_Classification, gene_protein, mutation_count,	region_size, tmb, VAF)

genomic_paired_df <- readr::read_tsv(genomic_paired_file, guess_max = 100000, show_col_types = FALSE) %>% 
  left_join(tmb_vaf_df, by = c("Kids_First_Biospecimen_ID"))

# Which patient samples don't have TMB?
#genomic_paired_df %>% 
#  filter(is.na(tmb)) %>% 
#  unique() %>% 
#  regulartable() %>%
#  fontsize(size = 12, part = "all")

# Vector to order timepoints
td_order <- c("Diagnosis", "Progressive", "Recurrence", "Deceased", "Second Malignancy", "Unavailable")

df <- genomic_paired_df %>% 
  filter(!is.na(tmb)) %>% 
  mutate(tumor_descriptor = factor(tumor_descriptor),
         tumor_descriptor = fct_relevel(tumor_descriptor, td_order))

# Let's count and remove any timepoints with less than 2 samples therein
#timepoint_n_df <- df %>% 
#  count(tumor_descriptor) %>% 
#  dplyr::mutate(tumor_descriptor_n = glue::glue("{tumor_descriptor} (N={n})")) %>% 
#  dplyr::rename(timepoint_n = n)

df_plot <- df #%>% 
#  left_join(timepoint_n_df, by = c("tumor_descriptor")) %>%
#  filter(!timepoint_n <= 2) # remove if total number of values per group is less than 2
``` 

# Type of non synonymous mutations per cancer group

```{r define-parameters-for-plots}
# Read color palette
palette_df <- readr::read_tsv(palette_file, guess_max = 100000, show_col_types = FALSE) 

# Define and order palette
palette <- palette_df$hex_codes
names(palette) <- palette_df$Variant_Classification

# Define label for plots
Alteration_type <- df_plot$Variant_Classification

# Define ylim
ylim <- max(df_plot$tmb)
```

# What type of alterations we observe per tumor descriptor in each timepoint model?

```{r plot-timepoints-models, fig.width = 20, fig.height = 15, fig.fullwidth = TRUE}
# Create bxp
print(df_plot %>% 
        ggplot2::ggplot() + 
        ggplot2::aes(x = tumor_descriptor, 
                     y = tmb,
                     color = Variant_Classification) +
        geom_boxplot() +
        ggplot2::scale_color_manual(values = palette, 
                                    breaks = sort(names(palette))) +
        facet_wrap(~timepoints_models) +
        theme_Publication() + 
        scale_y_continuous(limits = c(0, ylim)) +
        theme(legend.position = c(.95, .95),
              legend.justification = c("right", "bottom"),
              legend.box.just = "right",
              legend.margin = margin(6, 6, 6, 6)) +
        theme(axis.text.x = element_text(angle = 90)))

# Save the plot
ggsave(filename = "Alteration_type_timepoints_models.pdf", 
       path = plots_dir, 
       width = 20, 
       height = 15, 
       device = "pdf", 
       useDingbats = FALSE)
```

# TODO
- To subset per timepoint_model and count sample per tumor_descriptor (x axis).
- To fix legend position.


# What type of alterations we observe per tumor descriptor in each cancer group?

```{r plot-cancer-group, fig.width = 20, fig.height = 15, fig.fullwidth = TRUE}
# Create bxp
print(df_plot %>% 
        ggplot2::ggplot() + 
        ggplot2::aes(x = tumor_descriptor, 
                     y = tmb,
                     color = Variant_Classification) +
        geom_boxplot() +
        ggplot2::scale_color_manual(values = palette, 
                                    breaks = sort(names(palette))) +
        facet_wrap(~cancer_group) +
        theme_Publication() + 
        scale_y_continuous(limits = c(0, ylim)) +
        theme(legend.position = c(.95, .95),
              legend.justification = c("right", "bottom"),
              legend.box.just = "right",
              legend.margin = margin(6, 6, 6, 6)) +
        theme(axis.text.x = element_text(angle = 90)))

# Save the plot
ggsave(filename = "Alteration_type_cancer_group.pdf", 
       path = plots_dir, 
       width = 20, 
       height = 15, 
       device = "pdf", 
       useDingbats = FALSE)
```


# TODO
- To fix legend position.

```{r echo=TRUE}
sessionInfo()
```
